---
title: "scmr"
output:
  github_document:
    toc: true
    toc_depth: 3
    df_print: tibble
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# scmr: Spatially Clustered Multinomial Regression

`scmr` fits **spatially clustered multinomial regression** models for multiclass categorical responses observed over spatial units. The package was designed for reproducible simulation studies where the response is multinomial and the predictors may be high-dimensional and correlated. The main model combines:

1. local multinomial elastic-net regression models,
2. spatially constrained cluster membership,
3. a Potts-style spatial smoothness term, and
4. model-selection criteria including standard SCR-BIC and cluster-balanced CB-BIC.

The package also includes a non-spatial global penalized multinomial benchmark. Both models are called through one main function, `fit_scmr()`.

This repository intentionally contains **only synthetic-data simulation tools and examples**. It does not include private or restricted real data.

## Installation

Install the development version from GitHub:

```{r install, eval=FALSE}
# install.packages("remotes")
remotes::install_github("byususen/scmr")
```

Load the package:

```{r load-package, eval=FALSE}
library(scmr)
```

## Main user interface

The main function is:

```{r main-api, eval=FALSE}
fit_scmr(
  x,
  y,
  model = c("scmr", "global"),
  penalty = c("elastic_net", "lasso", "ridge"),
  G = NULL,
  unit_id = NULL,
  coords = NULL,
  alpha = NULL,
  alpha_grid = NULL,
  lambda = NULL,
  control = scmr_control(),
  initial_cluster = NULL,
  nfolds = 5,
  lambda_rule = c("lambda.1se", "lambda.min"),
  seed = 123
)
```

Use:

- `model = "global"` for a global penalized multinomial benchmark.
- `model = "scmr"` for spatially clustered multinomial regression.
- `penalty = "ridge"` for alpha = 0.
- `penalty = "lasso"` for alpha = 1.
- `penalty = "elastic_net"` for alpha in `(0, 1]`, either fixed by `alpha` or tuned over `alpha_grid`.

## Required data structure

`scmr` expects a prepared model matrix and aligned response/spatial objects. It does not require a raw survey or remote-sensing table.

### Global model

For `model = "global"`, the required objects are:

| Object | Type | Dimension | Description |
|---|---|---:|---|
| `x` | numeric matrix or sparse `Matrix` | `n_observations × p_predictors` | Predictor matrix |
| `y` | factor or character vector | `n_observations` | Multiclass response |

Example:

```{r data-global, eval=FALSE}
X <- model.matrix(~ phase_lag1 + month_factor + NDVI + EVI + RVI - 1, data = df)
y <- factor(df$phase)

fit_global <- fit_scmr(
  x = X,
  y = y,
  model = "global",
  penalty = "elastic_net",
  alpha_grid = c(0.6, 0.7, 0.8),
  lambda_rule = "lambda.1se",
  nfolds = 5,
  seed = 123
)
```

### SCMR model

For `model = "scmr"`, the required objects are:

| Object | Type | Dimension | Description |
|---|---|---:|---|
| `x` | numeric matrix or sparse `Matrix` | `n_observations × p_predictors` | Predictor matrix |
| `y` | factor or character vector | `n_observations` | Multiclass response |
| `G` | integer | length 1 | Number of spatial clusters |
| `unit_id` | character, factor, or integer vector | `n_observations` | Spatial unit identifier |
| `coords` | numeric matrix or data frame | `n_observations × 2` | Coordinates for each row |

The row order must be aligned:

```{r row-alignment, eval=FALSE}
x[i, ]      # predictors for row i
y[i]        # response for row i
unit_id[i]  # spatial unit of row i
coords[i, ] # coordinate of row i
```

Repeated rows per spatial unit are allowed. All observations with the same `unit_id` are assigned to the same spatial cluster. For panel data, `unit_id` should usually be the permanent spatial unit identifier.

Example:

```{r data-scmr, eval=FALSE}
df$phase <- factor(df$phase)
df$phase_lag1 <- factor(df$phase_lag1)
df$month_factor <- factor(df$month)

X <- model.matrix(
  ~ phase_lag1 + month_factor + NDVI + EVI + RVI - 1,
  data = df
)

y <- df$phase
unit_id <- df$id_subsegmen
coords <- as.matrix(df[, c("lati_mean", "long_mean")])

fit_spatial <- fit_scmr(
  x = X,
  y = y,
  model = "scmr",
  penalty = "elastic_net",
  G = 4,
  unit_id = unit_id,
  coords = coords,
  alpha_grid = c(0.6, 0.7, 0.8),
  control = scmr_control(
    phi = 1,
    k_neighbors = 5,
    min_units = 10,
    min_per_class = 5,
    max_iter = 10
  ),
  seed = 123
)
```

## Quick start with synthetic data

The package includes `simulate_scmr_data()` for fully reproducible examples. The simulated data are public and safe to share.

```{r quick-start-simulate, eval=FALSE}
library(scmr)

set.seed(1)

dat <- simulate_scmr_data(
  n_obs = 1200,
  n_new = 200,
  p = 25,
  active = 1:5,
  eta = 0.2,
  scenario = "clustered_balanced",
  seed = 1000
)

obs <- dat$observed
str(obs)
```

`obs` contains:

| Element | Description |
|---|---|
| `obs$X` | predictor matrix |
| `obs$y` | class labels |
| `obs$coords` | two-column spatial coordinates |
| `obs$unit` | spatial unit ID |
| `obs$region` | true simulation region, useful for diagnostics |
| `obs$beta` | true coefficients, simulation-only |
| `obs$prob` | true class probabilities, simulation-only |

For model fitting, only `X`, `y`, `coords`, and `unit` are required.

## Train/test split

Use `make_stratified_split()` to create a stratified test set. If `region` is supplied, the split is stratified by region and class.

```{r split, eval=FALSE}
test_idx <- make_stratified_split(
  y = obs$y,
  region = obs$region,
  test_prop = 0.30,
  seed = 2024
)

train_idx <- setdiff(seq_along(obs$y), test_idx)

X_train <- obs$X[train_idx, , drop = FALSE]
X_test  <- obs$X[test_idx, , drop = FALSE]

y_train <- obs$y[train_idx]
y_test  <- obs$y[test_idx]

unit_train <- obs$unit[train_idx]
unit_test  <- obs$unit[test_idx]

coords_train <- obs$coords[train_idx, , drop = FALSE]
coords_test  <- obs$coords[test_idx, , drop = FALSE]
```

## Fitting the global benchmark

### Elastic net

```{r fit-global-enet, eval=FALSE}
fit_global_enet <- fit_scmr(
  x = X_train,
  y = y_train,
  model = "global",
  penalty = "elastic_net",
  alpha_grid = c(0.6, 0.7, 0.8),
  lambda_rule = "lambda.1se",
  nfolds = 5,
  seed = 2024
)

fit_global_enet
```

### Lasso

```{r fit-global-lasso, eval=FALSE}
fit_global_lasso <- fit_scmr(
  x = X_train,
  y = y_train,
  model = "global",
  penalty = "lasso",
  lambda_rule = "lambda.1se",
  nfolds = 5,
  seed = 2024
)
```

### Ridge

```{r fit-global-ridge, eval=FALSE}
fit_global_ridge <- fit_scmr(
  x = X_train,
  y = y_train,
  model = "global",
  penalty = "ridge",
  lambda_rule = "lambda.1se",
  nfolds = 5,
  seed = 2024
)
```

### Fixed alpha and lambda

Use a fixed `lambda` when you want exact reproducibility without cross-validation.

```{r fit-global-fixed, eval=FALSE}
fit_global_fixed <- fit_scmr(
  x = X_train,
  y = y_train,
  model = "global",
  penalty = "elastic_net",
  alpha = 0.7,
  lambda = 0.01,
  seed = 2024
)
```

## Fitting SCMR

### Basic SCMR elastic-net model

```{r fit-scmr-basic, eval=FALSE}
ctrl <- scmr_control(
  phi = 1,
  k_neighbors = 5,
  weight_type = "binary",
  max_iter = 10,
  init_method = "kmeans",
  init_tries = 100,
  update_order = "random",
  min_units = 10,
  min_per_class = 5,
  tune_lambda = TRUE,
  tune_alpha_after_convergence = TRUE
)

fit_spatial_enet <- fit_scmr(
  x = X_train,
  y = y_train,
  model = "scmr",
  penalty = "elastic_net",
  G = 4,
  unit_id = unit_train,
  coords = coords_train,
  alpha_grid = c(0.6, 0.7, 0.8),
  control = ctrl,
  seed = 2024
)

fit_spatial_enet
```

### SCMR with lasso penalty

```{r fit-scmr-lasso, eval=FALSE}
fit_spatial_lasso <- fit_scmr(
  x = X_train,
  y = y_train,
  model = "scmr",
  penalty = "lasso",
  G = 4,
  unit_id = unit_train,
  coords = coords_train,
  control = ctrl,
  seed = 2024
)
```

### SCMR with ridge penalty

```{r fit-scmr-ridge, eval=FALSE}
fit_spatial_ridge <- fit_scmr(
  x = X_train,
  y = y_train,
  model = "scmr",
  penalty = "ridge",
  G = 4,
  unit_id = unit_train,
  coords = coords_train,
  control = ctrl,
  seed = 2024
)
```

### SCMR with fixed alpha and fixed lambda

If `lambda` is supplied, SCMR skips initial lambda tuning and uses the supplied value. For SCMR, `lambda` can be either a scalar or a vector of length `G`.

```{r fit-scmr-fixed-lambda, eval=FALSE}
fit_spatial_fixed <- fit_scmr(
  x = X_train,
  y = y_train,
  model = "scmr",
  penalty = "elastic_net",
  G = 4,
  unit_id = unit_train,
  coords = coords_train,
  alpha = 0.7,
  lambda = 0.01,
  control = scmr_control(
    phi = 1,
    k_neighbors = 5,
    min_units = 10,
    min_per_class = 5,
    max_iter = 10,
    tune_lambda = FALSE,
    tune_alpha_after_convergence = FALSE
  ),
  seed = 2024
)
```

Cluster-specific fixed lambda values:

```{r fit-scmr-cluster-lambda, eval=FALSE}
fit_spatial_cluster_lambda <- fit_scmr(
  x = X_train,
  y = y_train,
  model = "scmr",
  penalty = "elastic_net",
  G = 4,
  unit_id = unit_train,
  coords = coords_train,
  alpha = 0.7,
  lambda = c(0.02, 0.01, 0.01, 0.005),
  control = scmr_control(
    tune_lambda = FALSE,
    tune_alpha_after_convergence = FALSE
  ),
  seed = 2024
)
```

## Important control arguments

`scmr_control()` controls the spatial clustering and local model-fitting algorithm.

| Argument | Default | Meaning |
|---|---:|---|
| `phi` | `1` | Spatial smoothness coefficient |
| `k_neighbors` | `5` | Number of nearest neighbors for spatial weights |
| `weight_type` | `"binary"` | Spatial weight type, either `"binary"` or `"exp"` |
| `weight_bandwidth` | `NULL` | Bandwidth for exponential weights |
| `row_standardize_weights` | `FALSE` | Whether to row-standardize spatial weights |
| `max_iter` | `10` | Maximum SCMR membership-update iterations |
| `init_method` | `"kmeans"` | Initial cluster method, either `"kmeans"` or `"random"` |
| `init_tries` | `200` | Number of initialization attempts |
| `update_order` | `"random"` | Unit update order, either `"random"` or `"fixed"` |
| `min_units` | `10` | Minimum unique spatial units per cluster |
| `min_per_class` | `5` | Minimum observations per class per cluster |
| `tune_lambda` | `TRUE` | Whether to tune cluster-local lambda values |
| `tune_alpha_after_convergence` | `TRUE` | Whether to retune alpha after membership convergence |
| `holdout_val_prop` | `0.30` | Validation proportion for holdout tuning |
| `holdout_nlambda` | `60` | Number of lambda values in the holdout path |
| `local_lambda_factors` | `c(3, 2, 1, 1/2, 1/3)` | Local search multipliers around current lambda |
| `criterion_balance_gamma` | `1` | Balance exponent in CB-BIC |

Example with more explicit settings:

```{r control-detail, eval=FALSE}
ctrl <- scmr_control(
  phi = 1,
  k_neighbors = 5,
  weight_type = "binary",
  max_iter = 10,
  init_method = "kmeans",
  init_tries = 200,
  update_order = "random",
  min_units = 10,
  min_per_class = 5,
  holdout_val_prop = 0.30,
  holdout_nlambda = 60,
  lambda_1se_tol = 0.002,
  lambda_update_tol = 0.001,
  alpha_update_tol = 0.003,
  local_lambda_factors = c(3, 2, 1, 1/2, 1/3),
  final_alpha_delta = c(-0.1, 0, 0.1),
  tiny_movement_max_units = 5,
  tiny_movement_rate_tol = 0.001,
  tiny_movement_revert = TRUE,
  criterion_balance_gamma = 1
)
```

## Initial cluster labels

You can supply an initial partition with `initial_cluster`. The vector must have one label per **unique spatial unit**, not one label per observation row.

```{r initial-cluster, eval=FALSE}
unique_units <- levels(factor(unit_train))

initial_cluster <- sample(
  x = 1:4,
  size = length(unique_units),
  replace = TRUE
)

fit_init <- fit_scmr(
  x = X_train,
  y = y_train,
  model = "scmr",
  penalty = "elastic_net",
  G = 4,
  unit_id = unit_train,
  coords = coords_train,
  initial_cluster = initial_cluster,
  control = ctrl,
  seed = 2024
)
```

The supplied partition must satisfy the feasibility constraints defined by `min_units` and `min_per_class`.

## Prediction

### Global model prediction

```{r predict-global, eval=FALSE}
prob_global <- predict(fit_global_enet, newx = X_test, type = "prob")
pred_global <- predict(fit_global_enet, newx = X_test, type = "class")
```

### SCMR prediction

SCMR prediction needs the test predictors, test spatial unit IDs, and test coordinates.

```{r predict-scmr, eval=FALSE}
prob_scmr <- predict(
  fit_spatial_enet,
  newx = X_test,
  new_unit_id = unit_test,
  new_coords = coords_test,
  type = "prob"
)

pred_scmr <- predict(
  fit_spatial_enet,
  newx = X_test,
  new_unit_id = unit_test,
  new_coords = coords_test,
  type = "class"
)
```

To obtain cluster-membership probabilities for new units:

```{r predict-membership, eval=FALSE}
pp_test <- predict(
  fit_spatial_enet,
  newx = X_test,
  new_unit_id = unit_test,
  new_coords = coords_test,
  type = "membership"
)

head(pp_test)
```

## Evaluation metrics

Use `calc_overall_metrics()` and `calc_class_metrics()`.

```{r metrics, eval=FALSE}
classes <- levels(y_train)

overall_global <- calc_overall_metrics(
  y_true = y_test,
  y_pred = pred_global,
  prob = prob_global,
  classes = classes
)

overall_scmr <- calc_overall_metrics(
  y_true = y_test,
  y_pred = pred_scmr,
  prob = prob_scmr,
  classes = classes
)

rbind(
  global = overall_global,
  scmr = overall_scmr
)
```

Per-class metrics:

```{r class-metrics, eval=FALSE}
calc_class_metrics(
  y_true = y_test,
  y_pred = pred_scmr,
  classes = classes
)
```

## Model-selection criteria

Fitted objects store model criteria.

```{r criteria, eval=FALSE}
fit_global_enet$criteria
fit_spatial_enet$criteria
```

For SCMR, useful fields include:

| Field | Meaning |
|---|---|
| `CriterionSCR_BIC_original` | Standard SCR-BIC-type criterion |
| `CriterionV3_CB_BIC` or `CriterionV3_CB_uBIC` | Cluster-balanced criterion, depending on object version |
| `df_original_total` | Full parameter-count degrees of freedom |
| `df_effective_total` | Elastic-net effective degrees of freedom |
| `CBuBIC_Fit` | Cluster-balanced fit component |
| `CBuBIC_Penalty` | Cluster-balanced penalty component |
| `cluster_components` | Cluster-level diagnostic table |

Because object field names can change during development, inspect available fields with:

```{r inspect-criteria, eval=FALSE}
names(fit_spatial_enet$criteria)
```

## Reproducible simulation study

A complete simulation driver is included in:

```text
inst/scripts/01_run_simulation_study.R
```

You can run it from the package root:

```{r run-script, eval=FALSE}
source("inst/scripts/01_run_simulation_study.R")
```

A minimal simulation loop is shown below.

```{r simulation-loop, eval=FALSE}
scenarios <- c("global", "clustered_balanced", "clustered_imbalanced")
G_grid <- 1:6
n_repeats <- 5

results <- list()

for (rr in seq_len(n_repeats)) {
  for (scenario in scenarios) {
    dat <- simulate_scmr_data(
      n_obs = 1200,
      n_new = 200,
      p = 25,
      active = 1:5,
      eta = 0.2,
      scenario = scenario,
      seed = 1000 + rr
    )

    obs <- dat$observed

    test_idx <- make_stratified_split(
      y = obs$y,
      region = obs$region,
      test_prop = 0.30,
      seed = 2000 + rr
    )

    train_idx <- setdiff(seq_along(obs$y), test_idx)

    X_train <- obs$X[train_idx, , drop = FALSE]
    X_test  <- obs$X[test_idx, , drop = FALSE]
    y_train <- obs$y[train_idx]
    y_test  <- obs$y[test_idx]
    unit_train <- obs$unit[train_idx]
    unit_test  <- obs$unit[test_idx]
    coords_train <- obs$coords[train_idx, , drop = FALSE]
    coords_test  <- obs$coords[test_idx, , drop = FALSE]

    fit_g <- fit_scmr(
      X_train,
      y_train,
      model = "global",
      penalty = "elastic_net",
      alpha_grid = c(0.6, 0.7, 0.8),
      seed = 3000 + rr
    )

    pred_g <- predict(fit_g, X_test, type = "class")
    prob_g <- predict(fit_g, X_test, type = "prob")

    results[[length(results) + 1]] <- data.frame(
      Repeat = rr,
      Scenario = scenario,
      Model = "global",
      G = 1,
      calc_overall_metrics(y_test, pred_g, prob_g, levels(y_train))
    )

    for (G in G_grid) {
      fit_s <- fit_scmr(
        X_train,
        y_train,
        model = "scmr",
        penalty = "elastic_net",
        G = G,
        unit_id = unit_train,
        coords = coords_train,
        alpha_grid = c(0.6, 0.7, 0.8),
        control = scmr_control(
          phi = 1,
          k_neighbors = 5,
          min_units = 10,
          min_per_class = 5,
          max_iter = 10
        ),
        seed = 4000 + 100 * rr + G
      )

      pred_s <- predict(
        fit_s,
        X_test,
        new_unit_id = unit_test,
        new_coords = coords_test,
        type = "class"
      )

      prob_s <- predict(
        fit_s,
        X_test,
        new_unit_id = unit_test,
        new_coords = coords_test,
        type = "prob"
      )

      results[[length(results) + 1]] <- data.frame(
        Repeat = rr,
        Scenario = scenario,
        Model = "scmr",
        G = G,
        calc_overall_metrics(y_test, pred_s, prob_s, levels(y_train))
      )
    }
  }
}

results <- do.call(rbind, results)
head(results)
```

## Using your own data without publishing it

Your private data should be transformed locally into the required modeling objects:

```{r own-data-template, eval=FALSE}
# Private local table. Do not commit this file to GitHub.
df <- read.csv("path/to/private_data.csv")

# Example response and predictors.
df$y <- factor(df$y)
df$lag_y <- factor(df$lag_y)
df$month_factor <- factor(df$month)

X <- model.matrix(
  ~ lag_y + month_factor + predictor_1 + predictor_2 + predictor_3 - 1,
  data = df
)

y <- df$y
unit_id <- df$spatial_unit_id
coords <- as.matrix(df[, c("latitude", "longitude")])

fit <- fit_scmr(
  x = X,
  y = y,
  model = "scmr",
  penalty = "elastic_net",
  G = 4,
  unit_id = unit_id,
  coords = coords,
  control = scmr_control(min_units = 10, min_per_class = 5),
  seed = 123
)
```

Recommended repository practice:

- Commit scripts that create `X`, `y`, `unit_id`, and `coords`.
- Do **not** commit restricted raw data.
- Include a small synthetic example or anonymized toy dataset if needed.
- Add private file patterns to `.gitignore`, for example:

```text
private_data/
*.csv
*.rds
*.RData
```

If your reproducibility workflow needs real-data results but not real data, commit only aggregated tables that do not identify units or reveal restricted values.

## Saving outputs

The package returns R objects directly. A simple reproducibility script can save results as CSV/RDS.

```{r save-results, eval=FALSE}
saveRDS(fit_spatial_enet, "outputs/fit_spatial_enet.rds")
write.csv(overall_scmr, "outputs/test_metrics_scmr.csv", row.names = FALSE)
```

To save SCMR cluster membership for later analysis:

```{r save-membership, eval=FALSE}
pp_train <- fit_spatial_enet$PP_unit

membership_obj <- list(
  metadata = list(
    ModelType = "SCMR",
    Repeat = 1,
    FixedK = fit_spatial_enet$G
  ),
  unit = rownames(pp_train),
  cluster = as.integer(fit_spatial_enet$group_unit[rownames(pp_train)]),
  latitude = fit_spatial_enet$train_unit_coords[, 1],
  longitude = fit_spatial_enet$train_unit_coords[, 2],
  membership_matrix = pp_train
)

saveRDS(membership_obj, "outputs/scmr_membership_G4.rds", compress = "xz")
```

Read compact membership files back into long format:

```{r read-membership, eval=FALSE}
membership_long <- compact_membership_to_long("outputs/scmr_membership_G4.rds")
head(membership_long)
```

## Development workflow

Clone the repository, then run:

```{r dev-workflow, eval=FALSE}
install.packages(c("devtools", "roxygen2", "testthat"))

devtools::document()
devtools::check()
```

Build this README from `README.Rmd`:

```{r build-readme, eval=FALSE}
devtools::build_readme()
```

Run tests:

```{r test, eval=FALSE}
devtools::test()
```

## Citation

If you use this package in academic work, please cite the associated manuscript or dissertation chapter describing SCMR. A package citation file can be added later in `inst/CITATION`.

## License

See `LICENSE` for license details.
